A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-23 paper.Rmd
done with susns suggestions

Motivated by the advice of regression analysis experts, that residual plots as opposed to conventional tests are indispensible methods for assessing model fit, a human subjects experiment was conducted using visual inference. The experiment tested two primary departures from good residuals: non-linearity and heteroskedasticity.

The experiment found that conventional residual-based statistical tests are more sensitive to weak departures from model assumptions than visual tests as would be evaluated by humans. That is, a conventional test concludes there are problems with the model fit almost twice as often as a human would. They often reject when departures in the form of non-linearity and heteroskedasticity are not visibly different from null residual plots.

While it might be argued that the conventional tests are correctly detecting the small but real effects, it can also be considered that the conventional tests are rejecting when it is not necessary. Many of these rejections happen even when downstream analysis and results would not be significantly affected by the small departures from a good fit. The results from human evaluations provide a more practical solution, which reinforces the statements from regression experts that residual plots are an indispensible method for model diagnostics.

Now it is important to note that residual plots need to be delivered as a lineup, where it is embedded in a field of null plots. A residual plot may contain many visual features, but some are caused by the characteristics of the predictors and the randomness of the error, not by the violation of the model assumptions. These irrelevant visual features have a chance to be filtered out by participants with a comparison to null plots, resulting in more accurate reading. The lineup enables a careful calibration for reading structure in residual plots.

Human evaluation of residuals is expensive, though. It is time-consuming and laborious. This is possibly why residual plot analysis is often not done, in practice. However, with the emergence of effective computer vision this work could form the basis to provide automated residual plot reading. <!--The findings suggest the strong demand of graphical inspection in regression diagnostics, so developing an automatic visual inference system to evaluate lineups of residual plots is valuable. We plan to build a completed open-source system in an `R` package and provide a web interface such as a website for public to interact with. Details about this system will be discussed in our next paper.-->

The experiment also revealed some interesting results. For the most part, the visual test performed very similarly to the appropriate conventional test except that it has a power curve shifted in the less sensitive direction. Unlike the conventional tests, where one needs to specifically test for non-linearity or heteroskedasticity the visual test operated effectively across the range of departures from good residuals. If the fitted value distribution is not uniform, there is a small loss of power in the visual test. But quite a surprise to us was the big difference in power of the conventional test, which appear to be quite sensitive to how the predictors are sampled. Another surprising finding was that the direction of heteroskedasticity appears to affect the ability to visually detect it: both triangles being more difficult to detect than the butterfly, and a small difference in detection between left- and right-triangle.

>>>>

Regression analysis experts suggest that residual plots are indispensible methods for assessing model fit. We conducted a perceptual experiment using visual inference to assess this oft-repeated advice. The experiment tested two primary departures from good residuals: non-linearity and heteroskedasticity.

We found that conventional residual-based statistical tests are more sensitive to weak departures from model assumptions than visual tests. That is, a conventional test concludes there are problems with the model fit almost twice as often as a human. Conventional tests often reject when departures in the form of non-linearity and heteroskedasticity are not visibly different from null residual plots.

While it might be argued that the conventional tests are correctly detecting small but real effects, this can also be seen as the conventional tests are rejecting unnecessarily. Many of these rejections happen even when downstream analysis and results would not be significantly affected by the small departures from a good fit. The results from human evaluations provide a more practical solution, which reinforces the statements from regression experts that residual plots are an indispensible method for model diagnostics.

It is important to note that residual plots need to be delivered as a lineup, embedded in a field of null plots. A residual plot may contain many visual features, but some are caused by the characteristics of the predictors and the randomness of the error, not by the violation of the model assumptions. These irrelevant visual features have a chance to be filtered out by participants with a comparison to null plots, resulting in more accurate reading. The lineup enables a careful calibration for reading structure in residual plots.

Human evaluation of residuals is expensive, time-consuming and laborious. This is possibly why residual plot analysis is often not done in practice. However, with the emergence of effective computer vision, our work could lay a foundation for automated residual plot assessment. <!--The findings suggest the strong demand of graphical inspection in regression diagnostics, so developing an automatic visual inference system to evaluate lineups of residual plots is valuable. We plan to build a completed open-source system in an `R` package and provide a web interface such as a website for public to interact with. Details about this system will be discussed in our next paper.-->

The experiment also revealed some interesting results. For the most part, the visual test performed similarly to the appropriate conventional test with a shift in the power curve. Unlike conventional tests, where one needs to specifically test for non-linearity or heteroskedasticity the visual test operated effectively across the range of departures from good residuals. If the fitted value distribution is not uniform, there is a small loss of power in the visual test. Surprisingly there is a big difference in power of the conventional test across fitted value distributions. Another surprising finding was that the direction of heteroskedasticity appears to affect the ability to visually detect it: both triangles being more difficult to detect than the butterfly, and a small difference in detection between left- and right-triangle.

### 2023-04-23 paper.tex
done with susns suggestions

Motivated by the advice of regression analysis experts, that residual
plots as opposed to conventional tests are indispensible methods for
assessing model fit, a human subjects experiment was conducted using
visual inference. The experiment tested two primary departures from good
residuals: non-linearity and heteroskedasticity.

The experiment found that conventional residual-based statistical tests
are more sensitive to weak departures from model assumptions than visual
tests as would be evaluated by humans. That is, a conventional test
concludes there are problems with the model fit almost twice as often as
a human would. They often reject when departures in the form of
non-linearity and heteroskedasticity are not visibly different from null
residual plots.

While it might be argued that the conventional tests are correctly
detecting the small but real effects, it can also be considered that the
conventional tests are rejecting when it is not necessary. Many of these
rejections happen even when downstream analysis and results would not be
significantly affected by the small departures from a good fit. The
results from human evaluations provide a more practical solution, which
reinforces the statements from regression experts that residual plots
are an indispensible method for model diagnostics.

Now it is important to note that residual plots need to be delivered as
a lineup, where it is embedded in a field of null plots. A residual plot
may contain many visual features, but some are caused by the
characteristics of the predictors and the randomness of the error, not
by the violation of the model assumptions. These irrelevant visual
features have a chance to be filtered out by participants with a
comparison to null plots, resulting in more accurate reading. The lineup
enables a careful calibration for reading structure in residual plots.

Human evaluation of residuals is expensive, though. It is time-consuming
and laborious. This is possibly why residual plot analysis is often not
done, in practice. However, with the emergence of effective computer
vision this work could form the basis to provide automated residual plot
reading.

The experiment also revealed some interesting results. For the most
part, the visual test performed very similarly to the appropriate
conventional test except that it has a power curve shifted in the less
sensitive direction. Unlike the conventional tests, where one needs to
specifically test for non-linearity or heteroskedasticity the visual
test operated effectively across the range of departures from good
residuals. If the fitted value distribution is not uniform, there is a
small loss of power in the visual test. But quite a surprise to us was
the big difference in power of the conventional test, which appear to be
quite sensitive to how the predictors are sampled. Another surprising
finding was that the direction of heteroskedasticity appears to affect
the ability to visually detect it: both triangles being more difficult
to detect than the butterfly, and a small difference in detection
between left- and right-triangle.

>>>>

Regression analysis experts suggest that residual plots are
indispensible methods for assessing model fit. We conducted a perceptual
experiment using visual inference to assess this oft-repeated advice.
The experiment tested two primary departures from good residuals:
non-linearity and heteroskedasticity.

We found that conventional residual-based statistical tests are more
sensitive to weak departures from model assumptions than visual tests.
That is, a conventional test concludes there are problems with the model
fit almost twice as often as a human. Conventional tests often reject
when departures in the form of non-linearity and heteroskedasticity are
not visibly different from null residual plots.

While it might be argued that the conventional tests are correctly
detecting small but real effects, this can also be seen as the
conventional tests are rejecting unnecessarily. Many of these rejections
happen even when downstream analysis and results would not be
significantly affected by the small departures from a good fit. The
results from human evaluations provide a more practical solution, which
reinforces the statements from regression experts that residual plots
are an indispensible method for model diagnostics.

It is important to note that residual plots need to be delivered as a
lineup, embedded in a field of null plots. A residual plot may contain
many visual features, but some are caused by the characteristics of the
predictors and the randomness of the error, not by the violation of the
model assumptions. These irrelevant visual features have a chance to be
filtered out by participants with a comparison to null plots, resulting
in more accurate reading. The lineup enables a careful calibration for
reading structure in residual plots.

Human evaluation of residuals is expensive, time-consuming and
laborious. This is possibly why residual plot analysis is often not done
in practice. However, with the emergence of effective computer vision,
our work could lay a foundation for automated residual plot assessment.

The experiment also revealed some interesting results. For the most
part, the visual test performed similarly to the appropriate
conventional test with a shift in the power curve. Unlike conventional
tests, where one needs to specifically test for non-linearity or
heteroskedasticity the visual test operated effectively across the range
of departures from good residuals. If the fitted value distribution is
not uniform, there is a small loss of power in the visual test.
Surprisingly there is a big difference in power of the conventional test
across fitted value distributions. Another surprising finding was that
the direction of heteroskedasticity appears to affect the ability to
visually detect it: both triangles being more difficult to detect than
the butterfly, and a small difference in detection between left- and
right-triangle.
